Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-08-24 response.tex

\usepackage[parfill]{parskip}

\newcommand{\comment}[1]{\textit{#1}}
\newcommand{\response}[1]{#1}
\newcommand{\todo}[1]{{\color{red} #1}}

\begin{document}

\response{We thank the AE and two reviewers for their comments which dramatically improved the manuscript.}

Associate Editor comments:

@@

\comment{Perhaps a reference to the promise of allele specific expression analyses (Discussion section) for inferring upon gene-expression heterosis might be useful to point readers into other directions with this work (Bell et al., 2013; Wei and Wang, 2013).}

\todo{Add references to allele specific expression.}

\comment{I realize that Ji et al.’s (2014) work was intended for microarray data, or more specifically Gaussian likelihoods, but there has been some work that suggests that Gaussian likelihood specifications in MAANOVA may, oddly enough, have better properties for the analysis of RNA-seq data than more popular methods specifically intended for count data (edgeR and DESeq).  See Reeb and Steibel (2013) and also comments from one reviewer along these lines.  Since Ji et al.’s work also derives from ISU, have you assessed its performance relative to your proposed method on RNA-seq data?}

\response{We have included Ji et al.'s work as another alternative method, see Sections 3.2 and 3.3.}

@@

\comment{Page 3 Line 26.  There are several ways to construct a Poisson-gamma distribution, all leading to different specifications of the relationship between the mean and variance of a negative binomial.  I apologize for the admittedly vain self-citation but there is one specification, for example, that specifies the variance to be a linear function of the mean of a negative binomial…see Equation 2.12b in  Tempelman and Gianola (1996).  I guess I’m not sure why most RNA-seq data analysts gravitate to a specification where the variance is specified to be linear and quadratic function of the mean as you do here.}

\todo{??}

\comment{Page 3. Bottom.  Shouldn’t you at least provide a justification when one would use a normal prior and when one would or should use a Laplace prior?}

\response{We have assessed the use of normal distributions see Section 3.3 and also provided rationale for why Laplace distributions would be expected to perform better in this scenario at the end of Section 2.1.}

@@

\comment{Page 5, lines 37-47.  Since you’re doing MCMC, why not simply tabulate the number of times samples of $\mu_3$ exceeds either $\mu_1$ or $\mu_2$….and/or the number of times that $\mu_3$ is less than $\mu_1$ or $\mu_2$.  I suspect that what your test is really doing….in which case it might be useful to explicitly state this equivalence as such.}

\response{Yes, that is what the test is doing. We have now stated the case for LPH, i.e. ``$\delta_g < -|\alpha_g|$ (equivalently $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$).''

\comment{Page 6, top.  Is it possible then to assess your proposed method using Laplace versus normal priors to assess if that seems to be contributing to any differences between ShrinkBayes and your proposed method?  Maybe that is the real reason for the small improvements?...., and not the differences in sophistication of the constructed statistical tests?}

\response{We have assessed the use of normal distributions, see Section 3.3.}

@@

Reviewer \#1

\comment{The authors develop a hierarchical negative binomial model for drawing inferences from a composite null hypothesis, that tests gene expression heterosis. Simulations are performed to describe the performance of this new method along with a data analysis demonstrating an application to real data. This manuscript represents a well-conceived and executed methodological development in the statistical analysis of RNA-Seq gene expression data. However, the performance of the new method is underwhelming compared to existing methods, tempering my enthusiasm for the likely impact of this paper.}

\comment{(1) In the simulations, the authors should also compare the eBayes method to the method of Ji et al. (2014). Although developed for continuous microarray data, the authors could apply a simple transformation to the count data in order to make them continuous. I would be curious to know whether the eBayes method out-performs the method of Ji et al. (2014) on the transformed data.}

\response{We have assessed the approach of Ji et al. on transformed data, see Sections 3.2 and 3.3.}

### 2015-08-24 ebayes_rnaseq_heterosis.tex
prose for figs

We used the methods described at the beginning of Section \ref{s:ebayes} to obtain normalization factors $\hat{\gamma}$ and gene-specific parameter estimates $\hat{\theta}_g$ for all genes using the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} applied to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across the four replicates.  This analysis produced sample-specific normalization factors of $\hat{\gamma}=
(0.074, -0.059, -0.074, -0.014, -0.014, -0.124,  0.093,  0.063,  0.021, -0.037,  0.049,  0.021)$. The  gene-specific parameter estimates were treated as the true parameter values for the simulation study so that our simulated datasets mimicked the existing structure among the gene-variety means of the original maize data. If $\hat{\delta}_g<-|\hat{\alpha}_g|$ ($\hat{\delta}_g >|\hat{\alpha}_g|$)  the gene was considered to have LPH (HPH).

Using these parameters and normalization factors, we simulated data according to the negative binomial model in \eqref{e:data} independently for each gene. For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.  We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.  
% From Dan "Did we consider changing the number of zeros allowed when the number of reps is higher?  I like the idea of simplifying all this by just picking an average count threshold that his high enough to avoid problems.  I think that may be what others have done in the past."

@@

In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the overdominance context. For each method, we attempted to provide a measure of the strength of overdominance for each gene such that large values of this measure indicate support for heterosis.

\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we define a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.

\baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for overdominance.

@@

For the methods in Sections \ref{s:ebayes}, \ref{s:shrinkbayes}, and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data}, 
% From Dan, "why this citation?"
from those without heterosis \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Representative ROC curves for false positive rates below 0.1 for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
\label{f:roc}
\end{figure}
For this simulation, we can see that the approaches based most closely on the model in Section \ref{s:model}, i.e. eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.

Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} as described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect. Each line is a different simulation while the blue box indicates mean AUC (plus or minus one standard error).}
\label{f:auc}
\end{figure}
The pattern appears to hold, with the eBayes and \ShrinkBayes{} methods outperforming the other two. With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and our eBayes approach, but as the number of replicates increases our eBayes approach appears to improve relative to \ShrinkBayes{}.

>>>>

We used the methods described at the beginning of Section \ref{s:ebayes} to obtain normalization factors $\hat{\gamma}$ and gene-specific parameter estimates $\hat{\theta}_g$ for all genes using the \edgeR{} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} applied to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across the four replicates.  This analysis produced sample-specific normalization factors of $\hat{\gamma}=
(0.074, -0.059, -0.074, -0.014, -0.014, -0.124,  0.093,  0.063,  0.021, -0.037,  0.049,  0.021)$. The  gene-specific parameter estimates were treated as the true parameter values for the simulation study so that our simulated datasets mimicked the existing structure among the gene-variety means of the original maize data.

Using these parameters and normalization factors, we simulated data according to the negative binomial model in \eqref{e:data} independently for each gene. For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.  We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.

For a particular gene, the truth was determined via the estimated values for $\alpha_g$ and $\delta_g$. Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have overdominance. For many overdominance genes, the value of $|\hat{\delta}_g|$ was only slightly larger than $|\hat{\alpha}_g|$. Thus there are many genes in these simulated data sets that are considered to have overdominance, but whose signal in the simulations will be extremely small. Conversely, there are many non-overdominance genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many overdominance genes. Therefore, we expect it will be difficult to accurately identify overdominance genes, but believe this is indicative of real applications as well.

@@

In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the overdominance context. For each method, we attempted to provide a measure of the strength of overdominance for each gene such that large values of this measure indicate support for overdominance.

\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of overdominance, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we define a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no overdominance of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for overdominance.

\baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of overdominance for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for overdominance.

@@

For the methods in Sections \ref{s:ebayes}, \ref{s:shrinkbayes}, and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for overdominance. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with overdominance, as defined in Section \ref{s:sim_data}, 
% From Dan, "why this citation?"
from those without overdominance \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Example ROC curves for false positive rates below 0.1 for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
\label{f:roc}
\end{figure}
The ROC curves are modest, e.g. for a false positive rate of 5\%, the best performing methods only achieved a true positive rate of just over 15\%. This is consistent with our expectation discussed in Section \ref{s:sim_data} due to the high signal-to-noise ratio in these simulated data.

For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e. eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with overdominance from genes without improves.

Figure \ref{f:auc} provides the area under the ROC curve (AUC) below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves (AUC) below a false positive rate of 0.1 for 3 different replicates per variety for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} as described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect. Each line is a different simulation while the blue box indicates mean AUC (plus or minus one standard error).}
\label{f:auc}
\end{figure}
Similar to the single ROC curve, the eBayes and \ShrinkBayes{} methods appear to outperform the other method. With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases the eBayes approaches appear to improve relative to \ShrinkBayes{}. Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of overdominance is slightly different due to INLA not providing a full posterior.

There also appears to be a pattern of the eBayes approach with Laplace distributions for the half-parental differences and hybrid effect systematically performing better than the approach that uses normal distributions. We assume this is because the Laplace distributions are better approximations to the true underlying distribution for these parameters.
